---
title: "GLM Practice"
author: "Wendi White"
date: "11/22/2020"
output: html_document
---

github--  https://github.com/wendi-white/biol607_homework 

```{r}
knitr::opts_knit$set(root.dir = "/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data") #setwd for rest of file
```


```{r}
#libraries
library(ggplot2)
library(car)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(emmeans)
library(readr)
library(brms)
library(tidybayes)
library(ggdist)
library(bayesplot)
library(purrr)
library(modelr)
library(rsample)
library(boot)
```


1. Comparing Means
To start with, let’s warm up with a simple one-way ANOVA model. This example, from Whitlock and Schluter chapter 15 question 22 looks at the mass of lodgepole pinecones from different habitats.

1.1. Load and plot the data. Choose a plot that not only shows the raw data, but also the means and SE or CI of those means. +1 EC if Michael thinks it’s fancy.
```{r}
setwd("/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data")

pinecones <- read.csv("/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data/15q22LodgepolePineCones.csv")

ggplot(pinecones, 
       mapping=aes(x=habitat, y=conemass)) + #x hab y conemass
  stat_summary(color="red", size=1.3) + #plots the SE of the mean
  geom_boxplot(alpha=0.7) + #transparency of point
  theme_bw(base_size=17) #font size
```


1.2 Fit a model using least squares and evaluate all relevant assumptions. List them out as you test them. Can we use this model? If not, fix it. But if we can, no fix is needed!
```{r}
pines_lm <- lm(conemass~habitat, data=pinecones)
#assumptions
plot(pines_lm, which=1)  #residuals vs fitted values
plot(pines_lm, which=2) #qq plot 
shapiro.test(residuals(pines_lm)) #test for normality.we do have normal disk
plot(pines_lm, which=4) #cooks distances vs row labels
plot(pines_lm, which=5) #plot of residuals against leverages 
```


1.2 How much variation is explained by your model?

88.51% of variation in cone mass is explained by habitat

```{r}
summary(pines_lm) #summary of model
```



1.3 Show which means are different from each other. Are you correcting p-values? If so, how, and justify your choice.

Island absent vs island present and island absent vs mainland present habitats have means that are different from each other.

We don't really need to correct the p-value as we can see it doesn't really change any of our answers. AKA when we do add adjust= "none" it doesn't change our answer.

```{r}
pines_em <- emmeans(pines_lm, ~ habitat) #get least squares means for each habitat 
pines_em
#compare the diff habitats using Tukey test
pines_cont <- contrast(pines_em, method = "tukey") %>%
    plot()+ #plot it with a xint of 0 to see diff
  geom_vline(xintercept=0, color= "red")
pines_cont

?knitr::knit

#test with no adjustment and find very little difference but not enough diff to show significant diff between 
#island present and mainland present
pines_cont_adj <- contrast(pines_em, method = "tukey", adjust = "none") %>%
  plot()+ #plot it with a xint of 0 to see diff
  geom_vline(xintercept=0, color= "red")
pines_cont_adj
```



2. Comparing Means from Multiple Categories
In a study from Rogers et al. (2020) link, the authors performed an experiment where they moved panels that had been colonized by invertebrates on a dock to a nearby rocky jetty where predators could access panels. To separate out the effects of changes in abiotic environment versus predation, they performed a factorial experiment, either caging or not caging panels and placing them either on the side of a cinder block or hanging on a piece of PVC attached to the block where predators would have little access (but weren’t entirely stopped). They then looked at change in total cover of invertebrates. Using this old data file dug off of my hard drive, let’s see what they found.

2.1. Load and plot the data. We are interested in change in percent cover. Choose a plot that not only shows the raw data, but also the means and SE or CI of those means. +1 EC if Michael thinks it’s fancy.

```{r}
setwd("/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data")
inverts <- read_csv("/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data/fouling_transplant_data.csv")
inverts

inverts <- janitor::clean_names(inverts) #take out spaces and capitalization from col names
inverts

ggplot(inverts, #plot data to look at change in cover across treatments
       mapping=aes(x=treatment, y=change_in_cover,
                   color=position_on_block)) +
  stat_summary(color="black", size=0.8) + #plots the SE of the mean
  geom_point(alpha=0.7) + #tranparency of point
  theme_bw(base_size=17) #font size
```


2.2 Fit a model using likelihood and evaluate all relevant assumptions. Do you meet assumptions?

NOPE!! Specifically qq plot seems wonky, residuals vs fitted are pretty scattered, and Shapiro gives us a 0.01687 output which says we don't have a good model... we are rejecting the null that says we a normal distribution. 


```{r}
#use glm b/c looking at likelihood
inverts_glm <- glm(change_in_cover ~ caged*position_on_block, 
                   data = inverts,
                   family = gaussian(link="identity"))#create glm of our data that looks at the interaction of caged and position on block's influence on %cover

plot(inverts_glm, which=1)  #residuals vs fitted values
plot(inverts_glm, which=2) #qq plot 
shapiro.test(residuals(inverts_glm)) #test for normality 
plot(inverts_glm, which=4) #cooks distances vs row labels
```


2.3 If you answered yes to the above…. you are wrong. It doesn’t! Percentage data is weird. Difference in percentages can be ever weirder! There are three tried and true solutions here. But they MIGHT not all work.
      
      Incorporate initial cover as a covariate. This takes out that influence, and as such we’re looking at residuals of change. This sometimes, but not always, works.

```{r}
inverts_glm_int <- glm(change_in_cover ~ caged*position_on_block + initial_cover, 
                       data = inverts,
                       family = gaussian(link="identity")) #create glm of our data that looks at the interaction of caged and position on block's influence on %cover and also accounts for the influence that initial cover has on this interaction


plot(inverts_glm_int, which=1)  #residuals vs fitted values look more scattered
plot(inverts_glm_int, which=2) #qq plot looks much better
shapiro.test(residuals(inverts_glm_int)) #test for normality reject null
plot(inverts_glm_int, which=4) #cooks distances vs row labels looks good
```


      Divide change by initial cover to express change as percent change relative to initial cover.

```{r}
inverts_change_cov <- inverts %>%
  mutate(change_in_cover = (change_in_cover / initial_cover)) #mutate to make new col looking at % change in cover
inverts_change_cov

inverts_glm_change_cov <- glm(change_in_cover ~ caged*position_on_block, 
                              data = inverts_change_cov,
                              family = gaussian(link="identity"))#create lm of our transformed data that looks at the interaction of caged and position on block's influence on %cover

plot(inverts_glm_change_cov, which=1)  #residuals vs fitted values
plot(inverts_glm_change_cov, which=2) #qq plot better than original but not great
shapiro.test(residuals(inverts_glm_change_cov)) #test for normality and fail to reject null
plot(inverts_glm_change_cov, which=4) #cooks distances vs row labels. cooks distance is way smaller
```


      Calculate difference in logit cover (so, logit(initial cover) - logit(final cover)). Logit transformations linearize percent cover data, and are often all that is needed to work percent cover into a linear model. You can use car::logit() for this.
      
```{r}
inverts_change_logist <- inverts %>%
  mutate(change_logit_cover = car::logit(initial_cover) - car::logit(final_cover)) #use logit transformation to linearize percent cover data

inverts_glm_logit <- glm(change_logit_cover ~ caged*position_on_block, 
                         data = inverts_change_logist,
                         family = gaussian(link="identity")) #create lm of our transformed data that looks at the interaction of caged and position on block's influence on %cover

plot(inverts_glm_logit, which=1)  #residuals vs fitted values is better
plot(inverts_glm_logit, which=2) #qq plot is looking wonky
shapiro.test(residuals(inverts_glm_logit)) #test for normality reject null
plot(inverts_glm_logit, which=4) #cooks distances vs row labels. cooks looks smaller
```


      Try all three methods. Which one works so that you can produce valid inference?
 
I would argue that dividing change by initial cover to express change as percent change relative to initial cover is the best b/c the plots look good and the shaprio test gives us a higher p-value. Incorporating the initial cover as a covariate is also a very good option. When we look at cooks distance it looks like there's one data point in particular that is impacting this from being the better option. It looks good other than the p-value on the shaprio test is lower. 

2.4 Great! So, take us home! Using NHST with an alpha of 0.08 (why not), what does this fit model tell you about whether predation matters given how I have described the system? Feel free to replot the data or fit model results if helpful

The Anova comparison (for both option 1 and 2) tells us that caged and position on block both impact the number of inverts.  

```{r}
#method 1
Anova(inverts_glm_int) 

#method 2
Anova(inverts_glm_change_cov)
```



3. Comparing Means with Covariates
We will wrap up with a model mixing continuous and discrete variables. In this dataset from Scantlebury et al, the authors explored how caste and mass affected the energy level of naked mole rats.

3.1 OK, you know what you are about at this point. Load in the data, plot it, fit it, check assumptions. Use Bayes for this.
```{r}
setwd("/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data")
rats <- read_csv("/Users/wendiwhite/Desktop/Ecology/UMass Boston Masters/UMass Masters classes/bio stats 607/data/18e4MoleRatLayabouts.csv")
summary(rats)

ggplot(data= rats,#plot data of change in mass across energy and sort by caste
       mapping = aes(x=lnenergy,
                     y=lnmass,
                     color=caste)) + 
  geom_point(alpha=0.7) + #transparency of point
  theme_bw(base_size=17) #size of text

rat_brm <- brm(lnenergy~ lnmass + caste, #bayes model looking at the impacts of mass on energy while also considering caste
               family = gaussian(link="identity"),
               data=rats,
               chain= 3)

plot(rat_brm) #look at the bayes plot

rhat(rat_brm)%>%
  mcmc_rhat()#can see all values are close to 1

pp_check(rat_brm, "dens_overlay") #green is posteriors from fit
                #black is dist of length in bayes
                #looks like a good fit!


#checking our residuals
pp_check(rat_brm, "error_hist") #residuals look good!
pp_check(rat_brm, "error_scatter") #relationship 
        #between residuals and obs values.. see that it's  
        #generally linear so we're good!
pp_check(rat_brm, "error_scatter_avg") #takes avg error
          #over all posteriors

#no auto correlation 
mcmc_acf(rat_brm)
```


3.2 Examine whether there is an interaction or not using LOO cross-validation. Is a model with an interaction more predictive?

There is no clear interaction of caste and mass b/c loo_compare shows that rat_brm is the better model.

```{r}
rat_brm_int <- brm(lnenergy~ caste* lnmass, #bayes model looking at the impacts of caste, mass and the interaction of both on energy
               family = gaussian(link="identity"),
               data=rats,
               chain= 3)

rat_brm_loo <- loo(rat_brm) #leave one out inference for the rat bayes model
rat_brm_int_loo <- loo(rat_brm_int) #leave one out inference for the rat bayes model that includes the interaction

loo_compare(rat_brm_loo, rat_brm_int_loo) #compare to see which model is correct

#see that rat_brm is 0 which means we select this one. We don't pick model with interaction of caste and lnmass
```


3.3 Compare the two castes energy expenditure at the mean level of log mass. Are they different? How would you discuss your conclusions.

Our 95% confidence intervals are very diff. Our lower worker CI and upper lazy do not overlap which tells us that there is a very diff energy expenditure between lazy and worker.

```{r}
emmeans(rat_brm, ~caste, method="tukey") #compare means of bayes model looking at energy expenditure between lazy and worker

emmeans(rat_brm, ~caste) %>% #do it as a comparison between the two castes
  contrast(method="tukey")
```


3.4 Plot the fit model. Use tidybayes and ggdist with your model to show fit and credible intervals with the raw data points on top. modelr::data.grid() might help as well.

```{r}
rat_brm <- brm(lnenergy~ lnmass + caste, #bayes model looking at the impacts of mass on energy while also considering caste
               family = gaussian(link="identity"), #gaussian dist and identity for link
               data=rats,
               chain= 3)

rat_newdat <- modelr::data_grid(rats, #do sequence of 100 over range of lnmass per each caste
                                  caste = unique(caste),
                                  lnmass= seq_range(lnmass, n=100))

rat_predict <- predict(rat_brm, #look at the response and CI that we get from our bayes model  and add those predictions to the rat_newdat.
                         newdata= rat_newdat,
                         type= "response",
                         interval= "confidence")

rat_newdat <- rat_newdat %>%
  mutate(lnenergy= rat_predict) #add a new column with the CI

#find predictions of the data and create new data frame with the means
rat_newfit <- emmeans(rat_brm, specs = ~caste +lnmass, #take the means of the data set across 100 of the mass data points
                        at = list(lnmass = seq(4,5, length.out = 100))) %>%
  as_tibble() %>%
  mutate(lnenergy= emmean)


ggplot(data= rats,#take origional graph of data looking at how mass changes energy for diff castes
       aes(x = lnmass,
           y = lnenergy,
           color= caste)) +
  geom_point() + #plot the origional data points
  geom_ribbon(data= rat_newfit, #use geom ribbon to set the upper and lower CI
              aes(ymin= lower.HPD, ymax= upper.HPD, group=caste),
              alpha = 0.1, color = "lightgrey")+ #set CI as light grey and transparency of 0.1 
  theme_classic()
```







